---
title: "Team Bike"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r warning=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)
library(chron)
library(Rtsne)
NYC <- read_csv("~/Desktop/NYC-3.txt")
# Removing the extraneous variables
NYC <- NYC %>%
  select(-drct,-p01i,-skyc1,-skyc2,-skyc3,-skyc4,skyl1,-skyl2,-skyl2,-skyl3,-skyl4,-metar)
randomsample <- read_csv("~/Desktop/compstats/ma154-project24-teambike/final_project/randomsample.csv")
```


```{r}

# Parsing the CitiBikes Data 
NYC$valid <- parse_date_time(NYC$valid,orders = "ymd HMS")

 


# Trying to get hourly data - one observation per hour, instead of multiple observations
NYC <- NYC %>%
  filter(minute(valid)=="51")
NYC <- NYC %>%
  mutate(Month=month(valid)) 

# Cieling the Hour 
NYC$valid <- ceiling_date(NYC$valid, unit = "hour")

# Making Binary Variables for the Seasons
NYC <- NYC %>%
  mutate(summer=ifelse(Month=="6"|Month=="7"|Month=="8",1,0)) %>%
  mutate(spring=ifelse((Month=="3"|Month=="4"|Month=="5"),1,0)) %>%
  mutate(winter=ifelse((Month=="1"|Month=="2"|Month=="12"),1,0)) %>%
  mutate(fall=ifelse((Month=="9"|Month=="10"|Month=="11"),1,0))
  

# Trying to Find Day of the Week
NYC <- NYC %>%
  mutate(WeekDay=wday(as.Date(valid), label=TRUE)) %>%
  mutate(WorkingDay=ifelse((WeekDay=="Mon"|WeekDay=="Tues"|WeekDay=="Wed"|WeekDay=="Thurs"|WeekDay=="Fri"),"Yes","No")) 

# Trying to find the holidays - Not quite working. Need to do this manually!
NYC <- NYC %>%
  mutate(holiday=is.weekend(valid))



# So I'm joing the two df's based on the date and the hour. The uid identifier is the date and hour. 
head(NYC$valid)
# take a look, I was able to join the two data frames based on the date and hour. I used hourly data when the minute is 51 to maintain uniformity.  

colnames(NYC)[2] <- "starttime"
```

```{r}
# Parsing all the start times into one format
mdy <- mdy_hms(randomsample$starttime) 
ymd <- ymd_hms(randomsample$starttime) 
f1 <- mdy_hm(randomsample$starttime)
mdy[is.na(mdy)] <- ymd[is.na(mdy)] # some dates are ambiguous, here we give 
randomsample$starttime <- mdy
randomsample$starttime[is.na(randomsample$starttime)] <- f1[is.na(randomsample$starttime)]

# Parsing all the end times into one format 
mdy <- mdy_hms(randomsample$stoptime) 
ymd <- ymd_hms(randomsample$stoptime) 
f1 <- mdy_hm(randomsample$stoptime)
mdy[is.na(mdy)] <- ymd[is.na(mdy)] 
randomsample$stoptime <- mdy
randomsample$stoptime[is.na(randomsample$stoptime)] <- f1[is.na(randomsample$stoptime)]

# Flooring the Hour
randomsample$starttime <- floor_date(randomsample$starttime,unit = "hour")

# Joining the weather data and the random sample
combined <- inner_join(NYC,randomsample)

# Number of Stations Every Year
rs <- randomsample %>%
  mutate(year=year(starttime)) %>%
  group_by(year,usertype) %>%
  summarise(stations=n_distinct(start.station.id),rides=n())


# Plotting the expansion of Citi Bikes Across New York
ggplot(rs,aes(x=year,y=stations,fill=usertype))+geom_bar(stat="identity")+ylab("Number of Stations")+xlab("Station")+ggtitle("Expansion of Citi Bikes Across New York")


```

```{r}
# Exploratory Analysis -
combined1 <- combined%>% 
  mutate(year=year(starttime),month=month(starttime),monyear=substr(starttime,1,7)) %>%
  group_by(monyear,year)%>%
  summarise(usage=n())

combined1$year <- as.factor(combined1$year)

# Graph to See the Monthly Usage
ggplot(combined1,aes(x=monyear,y=usage,fill=year))+geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle = 90, hjust = 1)) +ggtitle("Bike Usage in Different Months (July 2013 to September 2017)")+ylab("Number of Bike Rides")+xlab("Month and Year")

# Wrangling the data
combined2 <-  combined%>% 
  mutate(year=year(starttime),week=week(starttime),monyear=substr(starttime,1,7)) %>%
  group_by(week,year)%>%
  summarise(usage=n(),MedianDuration=median(tripduration)) %>%
  mutate(wy=paste(week,"-",year)) %>%
  filter(year==2016) %>%
  arrange(year)

# Graph to See the Usage in 2016
ggplot(combined2,aes(x=week,y=usage,fill=MedianDuration))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle = 90, hjust = 1))+ggtitle("Weekly Bike Rides in 2016")+ylab("Usage")+xlab("Week")

# Wrangling 
combined3 <- combined%>% 
  mutate(year=year(starttime),month=month(starttime),monyear=substr(starttime,1,7)) %>%
  group_by(WeekDay)%>%
  summarise(usage=n(),MedianDuration=median(tripduration))
# Graphing to See the Week Day
ggplot(combined3,aes(x=WeekDay,y=usage,fill=MedianDuration))+geom_bar(stat = "identity")

combined4 <- combined%>% 
  mutate(year=year(starttime),month=month(starttime),hour=hour(starttime)) %>%
  filter(!is.na(usertype))%>%
  group_by(hour,gender,usertype)%>%
  summarise(usage=n(),MedianDuration=median(tripduration))
# The Number of Rides for every hour - Will make binary variables for TSNE Clustering then. 

ggplot(combined4,aes(x=hour,y=usage,fill=usertype))+geom_bar(stat="identity")+xlab("Hour of the Day")+ylab("Number of Bike Rides")

# Making Different Levels for the Time of the Day
combined.clustering <- combined4 %>%
  mutate(EarlyMorning=ifelse(hour=="3"|hour=="4"|hour=="5"|hour=="6",1,0)) %>%
  mutate(Commuting=ifelse((hour=="7"|hour=="8"|hour=="9"),1,0)) %>%
  mutate(DayTime=ifelse((hour=="10"|hour=="11"|hour=="12"|hour=="13"|hour=="14"|hour=="15"),1,0)) %>%
  mutate(Evening=ifelse((hour=="16"|hour=="17"|hour=="18"|hour=="19"|hour=="20"),1,0)) %>%
  mutate(Night=ifelse((hour=="21"|hour=="22"|hour=="23"|hour=="0"|hour=="1"|hour=="2"),1,0))
combined4




dim(combined4)
combined4 <- combined4 %>%
  filter(!is.na(usertype)) %>%
  mutate(usertype=ifelse(usertype=="Customer",1,0))
combined4

# run Rtsne with default parameters - tsne 
tsne <- Rtsne(as.matrix(combined4), check_duplicates = FALSE, pca = FALSE, perplexity=30, theta=0.5, dims=2)
cols <- rainbow(10)
plot(tsne$Y, t='n')
text(tsne$Y, labels=combined4[,5])

```


Rationale for Choosing the weather data variables:

The variables relating to the weather data chosen are air temperature in Fahrenheit(tmpf), Dew Point Temperature in Fahrenheit (dwpf), Relative Humidity in percentage (relh), Wind Speed in knots (sknt), Pressure altimeter in inches (alti), and visibility in miles (vsby). We chose these variables because each of these variables can have an impact on a biker's decision to choose or not to choose to bike and can significantly affect bike usage. We chose these variables to build a preliminary model and make predictions based off the model. The other variables like wind direction in degrees from north, sky level coverages, sky level altitudes, and wind gust had some missing observations. We believe that these variables don't affect bike usage much and not including them would not affect the model. 



